---
title: "Distributions and Their Uses"
author: "Kevin Glass"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

``` {r echo=FALSE, message=FALSE}
options(scipen = 999)
options(repr.plot.width=15, repr.plot.height=8)


EXAMPLE = FALSE

library(dplyr)
library(ggplot2)
library(gridExtra)

```

# Overview

The purpose of this project is two fold: demonstrate the validity of the Central Limit Theorem (CLT), and apply the statistical inference methods covered in the JHU Statistical Inference course. The first part of the project shows the convergence of large samples from an exponential distribution converge to the theoretical value predicted by the CLT. The second part of the project demonstrates fundamental concepts of statistical inference such as Student's t-test, confidence intervals, group intervals, etc.

# Preliminaries
**Exponential Distribution**

The exponential distribution is defined by the function

\[f(x) = \bigg\{
  \begin{array}{ c l }
    \lambda e^{-\lambda x} & \quad \textrm{if } x \geq 0 \\
    0                 & \quad \textrm{if } x < 0
  \end{array}\]
  
where the rate parameter $\lambda$ is the mean number of events over a given time period^[https://en.wikipedia.org/wiki/Exponential_distribution]. The of mean and standard deviations of the exponential distribution are $1/\lambda$ and the variation is $1/\lambda^2$. Derivations for the mean^[https://statproofbook.github.io/P/exp-mean.html] and variance^[https://statproofbook.github.io/P/exp-var.html] can be found in \textbf{The Book of Statistical Proofs}. 

**Central Limit Theorem**

The Central Limit Theorem (CLT) says the mean of sample means of size n will approach
the population mean as $n \to \infty$. It also states that the standard deviation of the sample means $\sigma/\sqrt n$ will approach zero as $n \to \infty$ where $\sigma$ is the population mean. 

# Project 6, part 1: Simulating the CLT using the Exponential Distribution
The goal of this section is to demonstrate the validity of the Central Limit Theorem with respect the exponential distribution. It entails performing several simulations of an exponential distribution to demonstrate the convergence of sample means and the standard error to their theoretical values. It will also display some graphical results to visualize the convergence of the output



<!-- To do this, the code will generate random samples of varying sizes from an exponential distribution, and each sample size will have forty replicates. All samples are taken from an exponential distribution with a fixed rate constant $\lambda = 0.2$. -->

<!-- The code will find the average value of each sample and find the average value the forty replicates which is the mean of the sample means. It will also find the standard deviation of the sample means.  -->

<!-- The final analysis generated histograms using one replicate from some select samples. -->





<!-- the average of the sample means approaches the population mean.  -->


<!-- Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential samples. You should -->

<!-- Show the sample mean and compare it to the theoretical mean of the distribution. -->

<!-- Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution. -->

<!-- Show that the distribution is approximately normal. -->

<!-- In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials. -->


The previous section showed that the mean and variance of the exponential function are $1/\lambda$ and $1/\lambda^2$ respectively. This means the Central Limit Theorem 

```{r problem1, echo=FALSE}

# Requirements
# 1. Show sample mean and compare to theoretical mean
# 2. Show how variable the sample is (via variance) and compare it to the 
#    theoretical variance of the distribution.
# 3. Show that the distribution is approximately normal.
#    1. focus on the difference between the distribution of a large collection 
#       of random exponentials and the distribution of a large collection of 
#       averages of 40 exponentials.

# You will investigate the distribution of averages of 40 exponentials. 
# Note that you will need to do a thousand simulations.

if (!EXAMPLE)
{
  nSamples     <- 40                # required
  sampleSet    <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000, 2000, 3000)#, 4000, 5000, 6000, 7000, 8000, 9000, 10000, 20000, 30000, 40000, 50000, 60000, 70000, 80000, 90000, 100000)     # required

  lambda              <- 0.2           # required
  mu                  <- 1/lambda
  sigma               <- 1/lambda
  var                 <- 1/lambda^2
  sampleMeans         <- list()
  simulationSamples   <- data.frame()
  sampleDistribution  <- list()
  
  for (i in 1:length(sampleSet)) {
    exponentSampleMatrix <- matrix(rexp(sampleSet[i]*nSamples, lambda), 
                                nrow = sampleSet[i], ncol = nSamples)
    
    sampleDistribution[[i]] <- exponentSampleMatrix[, 1]
    # write.csv(exponentSampleMatrix, paste0("./expt", i, ".csv"),  row.names=TRUE)
    
    sampleMeans[[i]]  <- colMeans(exponentSampleMatrix) 
    simulationSamples    <- rbind(simulationSamples, list(sampleSet[i], 
                                                    mean(sampleMeans[[i]]), 
                                                    sd(sampleMeans[[i]])))
  }

  names(simulationSamples) <- c("size", "mean", "st_dev")
}

```

## Sample Mean
The sample mean is the center of mass of the observed data. 

```{r smGraph, echo = FALSE, warning = FALSE, fig.dim = c(4, 3)}
graphTics    <- c(5, 10, 100, 1000, 10000, 100000)     # required

if (!EXAMPLE) {
  ggplot(simulationSamples, aes(x=size)) +
    geom_line(aes(y=mean), color="blue") +
    geom_line(aes(y=st_dev), color="red") +
    geom_hline(yintercept = 5) +
    scale_x_log10(breaks = graphTics, limits = c(sampleSet[1], sampleSet[length(sampleSet)]+1)) +
    scale_y_continuous(limits = c( -0.1, 7.0 ))
}

```


```{r echo = FALSE, warning = FALSE}

if (!EXAMPLE) {
  
  # print(sampleMeans)
  # colSum <- sum(sampleMeans[[1]])
  # print(sampleMeans[[1]])
  # print(colSum)
  # print(sampleMeans[[1]]/colSum)
  # print(sampleMeans[[1]]/40)

  comparisonSet <- c(1, 3, 10)
  bw <- c(0.5, 0.5, 0.5, 0.5)
  p <- list()
  i  <- 1
  for (i in 1:length(comparisonSet)) {
    frame  <- data.frame(pdf = sampleMeans[[comparisonSet[[i]]]])

    p[[i]] <- ggplot(frame, aes(x = pdf)) +
      geom_histogram(binwidth = bw[i], fill="#000000", color="#e9ecef") +
    stat_function(fun = function (x)  dnorm(x, mean =  mu, sd = sigma/sqrt(nSamples))*25, color="red") +
      scale_x_continuous(limits = c(-0.1, 10.0)) +
      scale_y_continuous(limits = c(-0.1, 40.0))
  }
  frame  <- data.frame(pdf = sampleMeans[[19]])
  p[[4]] <- ggplot(frame, aes(x = pdf)) +
    geom_histogram(binwidth = bw[i], fill="#000000", color="#e9ecef") +
  stat_function(fun = function (x)  dnorm(x, mean =  mu, sd = sigma/sqrt(40))*25, color="green") +
    scale_x_continuous(limits = c(-0.1, 10.0)) +
    scale_y_continuous(limits = c(-0.1, 40.0))

  grid.arrange(p[[1]], p[[2]], p[[3]], p[[4]], nrow = 2, ncol = 2)
}


```

## Part two

```{r echo=FALSE, warning = FALSE}
# Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.
# Load the ToothGrowth data and perform some basic exploratory data analyses
# Provide a basic summary of the data.
# Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
# State your conclusions and the assumptions needed for your conclusions.

# The response is the length of odontoblasts (cells responsible for tooth 
# growth) in 60 guinea pigs. Each animal received one of three dose levels of 
# vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice 
# or ascorbic acid (a form of vitamin C and coded as VC).
# https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/ToothGrowth

doses    = c(0.5, 1.0, 2.0) # mg/day
delivery = c("OJ", "VC")    # delivery of VC, OJ - Orange Juice, VC - Ascorbic Acid

nPigs    <- 60
nLength  <- nPigs/(length(doses)*length(delivery))
nBins    <- nPigs/nLength

# print(ToothGrowth[1 : nLength,])
# print(nLength)
# print(ToothGrowth[(nLength + 1) : (2*nLength),])
# print(2*nLength)
# print(ToothGrowth[(2*nLength + 1) : (3*nLength),])
# print(3*nLength)
# print(ToothGrowth[(3*nLength + 1) : (4*nLength),])
# print(4*nLength)
# print(ToothGrowth[(4*nLength + 1) : (5*nLength),])
# print(5*nLength)
# print(ToothGrowth[(5*nLength + 1) : (6*nLength),])
# print(6*nLength)


# 
# ojGroup[[1]] <- ToothGrowth %>% filter(supp == "OJ" & dose == 0.5)
# ojGroup[[2]] <- ToothGrowth %>% filter(supp == "OJ" & dose == 1.0)
# ojGroup[[3]] <- ToothGrowth %>% filter(supp == "OJ" & dose == 2.0)
# aaGroup[[1]] <- ToothGrowth %>% filter(supp == "VC" & dose == 0.5)
# aaGroup[[2]] <- ToothGrowth %>% filter(supp == "VC" & dose == 1.0)
# aaGroup[[3]] <- ToothGrowth %>% filter(supp == "VC" & dose == 2.0)
bw <- c(3, 3, 4, 3, 3, 4)
p <- list()
if (!EXAMPLE) {
  for (i in 1:nBins){
    # data <-
    p[[i]] <- ggplot(ToothGrowth[((i-1)*nLength + 1) : (i*nLength),], aes(x=len, height = 100 , width = 3)) +
      geom_histogram( binwidth = bw[i], fill="#000000", color="#e9ecef") +
      scale_x_continuous(limits = c( 0, 35 ))
    # + 
    # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  }
 
}
grid.arrange(p[[1]], p[[4]], p[[2]], p[[5]], p[[3]], p[[6]], ncol = 2, nrow=3)
# grid.arrange(p4, p5, p6, ncol=3)



# print(ojGroup)
# print(aaGroup)

```


# Code

## Section one


```{r codeP1, echo=FALSE}

if (!EXAMPLE)
{
  nSamples     <- 40                # required
  sampleSet    <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 200, 300, 400, 
                    500, 600, 700, 800, 900, 1000, 2000, 3000)

  lambda           <- 0.2           # required
  mu               <- 1/lambda
  sigma            <- 1/lambda
  var              <- 1/lambda^2
  sampleMeans      <- list()
  sampleDistribution   <- list()        # accumulates data for 
  simulationSamples   <- data.frame()  # accumulates data for 
  
  for (i in 1:length(sampleSet)) {
    # fill the exponentsSet matrix with nSamples*sampleSet[i] values
    # transform matrix to N(0,1)
    exponentSampleMatrix <- matrix(rexp(sampleSet[i]*nSamples, lambda), 
                                nrow = sampleSet[i], ncol = nSamples)
    
    sampleDistribution[[i]] <- exponentSampleMatrix[, 1]
    # write.csv(exponentSampleMatrix, paste0("./expt", i, ".csv"),  row.names=TRUE)
    
    sampleMeans[[i]]  <- colMeans(exponentSampleMatrix)               # list of means for each matrix column
    simulationSamples    <- rbind(simulationSamples, list(sampleSet[i], 
                                                    mean(sampleMeans[[i]]), 
                                                    sd(sampleMeans[[i]])))
  }

  names(simulationSamples) <- c("size", "mean", "st_dev")
}

```

## Sample Mean
The sample mean is the center of mass of the observed data. 

```{r codeSMGraph, echo = FALSE, warning = FALSE, results="hide", fig.keep = "none"}
graphTics    <- c(5, 10, 100, 1000, 10000, 100000)     # required

if (!EXAMPLE) {
  ggplot(simulationSamples, aes(x=size)) +
    geom_line(aes(y=mean), color="blue") +
    geom_line(aes(y=st_dev), color="red") +
    geom_hline(yintercept = 5) +
    scale_x_log10(breaks = graphTics, limits = c(sampleSet[1], sampleSet[length(sampleSet)]+1)) +
    scale_y_continuous(limits = c( -0.1, 7.0 ))
}

```


```{r echo = FALSE, warning = FALSE, results="hide", fig.keep = "none"}

if (!EXAMPLE) {
  
  # print(sampleMeans)
  # colSum <- sum(sampleMeans[[1]])
  # print(sampleMeans[[1]])
  # print(colSum)
  # print(sampleMeans[[1]]/colSum)
  # print(sampleMeans[[1]]/40)

  comparisonSet <- c(1, 3, 10)
  bw <- c(0.5, 0.5, 0.5, 0.5)
  p <- list()
  i  <- 1
  for (i in 1:length(comparisonSet)) {
    frame  <- data.frame(pdf = sampleMeans[[comparisonSet[[i]]]])

    p[[i]] <- ggplot(frame, aes(x = pdf)) +
      geom_histogram(binwidth = bw[i], fill="#000000", color="#e9ecef") +
    stat_function(fun = function (x)  dnorm(x, mean =  mu, sd = sigma/sqrt(nSamples))*25, color="red") +
      scale_x_continuous(limits = c(-0.1, 10.0)) +
      scale_y_continuous(limits = c(-0.1, 40.0))
  }
  frame  <- data.frame(pdf = sampleMeans[[19]])
  p[[4]] <- ggplot(frame, aes(x = pdf)) +
    geom_histogram(binwidth = bw[i], fill="#000000", color="#e9ecef") +
  stat_function(fun = function (x)  dnorm(x, mean =  mu, sd = sigma/sqrt(40))*25, color="green") +
    scale_x_continuous(limits = c(-0.1, 10.0)) +
    scale_y_continuous(limits = c(-0.1, 40.0))

  grid.arrange(p[[1]], p[[2]], p[[3]], p[[4]], nrow = 2, ncol = 2)
}


```


