---
title: "Test"
author: "Kevin Glass"
date: "`r Sys.Date()`"
output: pdf_document
---

---
title           : "Distributions and Their Uses"
author          : "Kevin Glass"
date            : "`r Sys.Date()`"
output          : pdf_document

header-includes :
  - \usepackage{wrapfig}
  - \usepackage{caption}
  - \captionsetup[figure]{font = footnotesize}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data setup

\begin{verbatim}

        #                                 sampleRepetitions
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+ 
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+  40
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
        # +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+

 # Dist   ------------------------------------------------------------------------------> 10
 # Dist   ------------------------------------------------------------------------------> 20

\end{verbatim}


## Constants

```{r constants, echo=FALSE, message=FALSE, results = 'hide'}
options(scipen = 999)
options(digits = 3)

# Verify the simulation data is formatted as a list of n matrices
VERIFY_SOURCE_FILE = FALSE    # store to file
VERIFY_SOURCE_SCR  = FALSE    # print to screen
RUN_SIM_VERIFY     = FALSE    # run verification tests

# Verify the mean column data is stored as a list of lists.
VERIFY_MEAN_DISTRIB_FILE  = FALSE
VERIFY_MEAN_DISTRIB_SCR   = FALSE
RUN_VERIFY_MEAN_DISTRIB   = FALSE

# Verify the simulation data 
VERIFY_SAMPLE_STATS_FILE  = FALSE
VERIFY_SAMPLE_STATS_SCR   = FALSE
RUN_VERIFY_STATS          = FALSE

# Verify the mean distribution images 
VERIFY_MEAN_DIST_IMG      = TRUE
RUN_MEAN_DIST_IMG         = TRUE

library(dplyr)
library(ggplot2)
library(gridExtra)
library(knitr)
library(kableExtra)
library(ggpubr)
library(matrixStats)

set.seed(101)

```

## Simulation Data
### generateSimData

```{r generateSimData, echo=FALSE}

generateSimData <- function(sampleRepetition, sampleSize, lambda)
{
  if(RUN_SIM_VERIFY & VERIFY_MEAN_DISTRIB_SCR & TRUE)
  {
    print("generateSimData: verify inputs")
    print(paste0("sampleRepetitions = ", sampleRepetition))
    print(paste0("sampleSize   = ", sampleSize))
    print(paste0("lambda       = ", lambda))
  }

  # ----------------------------------------------------------------------------------
  simData <- list()

  for (i in 1:length(sampleRepetition)) {
    simData[[i]] <- matrix(rexp(sampleRepetition[i]*sampleSize, lambda), 
                                   nrow = sampleSize, ncol = sampleRepetition[i])
  }
  # ----------------------------------------------------------------------------------

  if(RUN_SIM_VERIFY & VERIFY_SOURCE_FILE & FALSE)
  {
    print("generateSimData: store in simData_#.csv")
    filename <- paste0("simData_", sampleRepetition[i], ".csv")
    write.csv(sampleMatrix, filename, row.names=FALSE)
  }

  if(RUN_SIM_VERIFY & VERIFY_SOURCE_SCR & TRUE)
  {
    print("generateSimData: ")
    print(simData)
  }

  return (simData)
}

if(RUN_SIM_VERIFY & TRUE)
{
  lambda       <- 0.2
  sampleSize   <- 40
  sampleRepetition  <- c(10, 20, 30, 40)
  
  simData <- generateSimData(sampleRepetition, sampleSize, lambda)

  print("RUN_SIM_VERIFY: simData")
  print(simData)
}

```

### calcMeanDistributions

```{r calcMeanDistributions, echo=FALSE}

calcMeanDistributions <- function(sampleRepetition, simData)
{
  if(RUN_VERIFY_MEAN_DISTRIB & VERIFY_MEAN_DISTRIB_SCR & TRUE)
  {
    print("generateSimData: verify inputs")
    print(paste0("sampleRepetitions = ", sampleRepetition))
    # print("simData = ")
    # print(simData)
    print("Test sim data separately")
  }

  # ----------------------------------------------------------------------------------
  meanDistributions <- list()
  
  for (i in 1:length(sampleRepetition)) {
    meanDistributions[[i]]   <- colMeans(simData[[i]])
  }
  # ----------------------------------------------------------------------------------
 
  return (meanDistributions)
}

if(RUN_VERIFY_MEAN_DISTRIB & TRUE)
{
  print("RUN_VERIFY_MEAN_DISTRIB: meanDistributions")
  lambda       <- 0.2
  sampleSize   <- 40
  sampleRepetition  <- c(10, 20, 30, 40)
  
  simData           <- generateSimData(sampleRepetition, sampleSize, lambda)
  meanDistributions <- calcMeanDistributions(sampleRepetition, simData)

  print("RUN_VERIFY_MEAN_DISTRIB: meanDistributions")
  print(meanDistributions)
}

```

## Data Analysis

### getSampleMeansStats

```{r getSampleMeansStats, echo=FALSE}

getSampleMeansStats <- function (sampleRepetition, meanDistributions, mu, var)
{
  if(RUN_VERIFY_STATS & VERIFY_SAMPLE_STATS_SCR & TRUE)
  {
    print("getSampleMeansStats: verify inputs")
    print(paste0("sampleRepetitions = ", sampleRepetition))
    print(paste0("mu           = ", mu))
    print(paste0("var          = ", var))
    # print("meanDistributions = ")
    # print(meanDistributions)
    print("Test meanDistributions separately")
  }

  # -----------------------------------------------------------------------------
  sampleStats <- data.frame(N_obs = integer(), mu = double(), muErr = double(), 
                            variation = double(), varErr = double())  
  for (i in 1:length(sampleRepetition)) {
    xbar     <- mean(meanDistributions[[i]])
    variance <- var(meanDistributions[[i]])

    sampleStats  <- rbind(sampleStats, 
                           c(sampleRepetition[i], xbar, abs((mu - xbar)/xbar),
                             variance, abs((var - variance)/variance)))
  }
  colnames(sampleStats) <- c("NRep", "Mean", "AbsErr_Mean", "Variance", "AbsEr_Var")
  # -----------------------------------------------------------------------------
  
  return (sampleStats)
}

if(RUN_VERIFY_STATS & TRUE)
{
  lambda       <- 0.2
  sampleSize   <- 40
  mu           <- 1/lambda
  var          <- mu^2/sqrt(sampleSize)
  sampleRepetition  <- c(10, 20, 30, 40)

  simData           <- generateSimData(sampleRepetition, sampleSize, lambda)
  meanDistributions <- calcMeanDistributions(sampleRepetition, simData)
  sampleMeansStats  <- getSampleMeansStats(sampleRepetition, meanDistributions, mu, var)
  print("RUN_VERIFY_STATS: sampleMeansStats")
  print(sampleMeansStats)
}

```


### imageDistribution

<!-- # ```{r echo = FALSE, warning = FALSE, results = 'hide', fig.dim = c(4, 3)} -->
```{r imageDistribution, echo = FALSE, warning = FALSE, fig.dim = c(4, 3)}

imageMeanDistribution <- function(sampleRepetition, sampleSize, meanDistributions, binRange, binName)
{
  if(RUN_MEAN_DIST_IMG & VERIFY_MEAN_DIST_IMG & TRUE)
  {
    print("imageMeanDistribution: verify inputs")
    print(paste0("sampleRepetitions = ", sampleRepetition))
    print(paste0("binRange          = ", binRange))
    print(paste0("binName           = ", binName))
    print(paste0("binName           = ", class(binName)))
    print(paste0("sampleSize        = ", sampleSize))
    print("meanDistributions = ")
    print(meanDistributions)
    # print("Test meanDistributions separately")
  }

  # -----------------------------------------------------------------------------
  label <- c("a", "b","c","d")
  pltList <- list()

  for (i in 1:length(sampleRepetition)) {
    counts <- binCounts(x = meanDistributions[[i]], 
                             idxs = NULL, 
                             bx   = binRange, 
                             nbins = 0)
    
    frame  <- data.frame(pdf = counts, binName = binName)
    colnames(frame) <- c("pdf", "binName")
    # frame  <- cbind(frame, binName)
    
    if(RUN_MEAN_DIST_IMG & VERIFY_MEAN_DIST_IMG & TRUE)
    {
      print(paste0("imageMeanDistribution: counts: ", counts))
      print(paste0("imageMeanDistribution: binRange: ", binRange))
      print(paste0("imageMeanDistribution: binRange: ", binName))
      print("imageMeanDistribution: frame")
      print(frame)
    }

    # pltList[i] <- ggplot(frame, aes(x = pdf)) +
    #   geom_histogram(binwidth = 0.5, fill="#000000", color="#e9ecef") #+
    #   # stat_function(fun = function (x)
    #   #   dnorm(x, mean =  mu, sd = sigma/sqrt(sampleSize))*25,
    #   #   color = "green") +
    #   # ggtitle(paste0("Sample Mean with n = ", sampleRepetition[i])) +
    #   # scale_x_continuous(limits = c(-1, 10.0), "Mean") +
    #   # scale_y_continuous(limits = c(-0.1, 20), "Count")
    # plot(pltList[[1]])
    
  }

  # plt <- ggarrange(pltList[[1]], pltList[[2]], pltList[[3]], pltList[[4]],
  #         labels = label,
  #         ncol = 2, nrow = 2)
  # 
  # ggsave(plot = plt, width = 6.0, height = 4.0, dpi = 300,
  #        filename = "distribution.png")
  # -----------------------------------------------------------------------------

}

if(RUN_MEAN_DIST_IMG & TRUE)
{
  lambda       <- 0.2
  sampleSize   <- 40
  # mu           <- 1/lambda
  # var          <- mu^2/sqrt(sampleSize)
  sampleRepetition  <- c(10, 20, 30, 40)
  binRange          <- seq(0.0, 10.0, by = 0.5)
  binName           <- list()
  for (i in 1:length(binRange) - 1) {
    binName <- append(binName, paste0("[", binRange[i], ", ", binRange[i+1], ")"))
  }
  
  simData           <- generateSimData(sampleRepetition, sampleSize, lambda)
  meanDistributions <- calcMeanDistributions(sampleRepetition, simData)
  imageMeanDistribution(sampleRepetition, sampleSize, meanDistributions, binRange, binName)
  print("RUN_MEAN_DIST_IMG: check image")
}


```

\begin{wrapfigure}{r}{0.3\textwidth}
  \centering
    \includegraphics[width=\linewidth]{distribution.png}
\end{wrapfigure}



### Teeth data

```{r teethData, echo=FALSE}
# 
# doses    = c(0.5, 1.0, 2.0) # mg/day
# delivery = c("OJ", "AA")    # delivery of AA, OJ - Orange Juice, AA - Ascorbic Acid
# 
# nPigs    <- 60
# nLength  <- nPigs/(length(doses)*length(delivery))
# nBins    <- nPigs/nLength
# bw <- c(3, 3, 4, 3, 3, 4)
# p <- list()
# 
#   doseFactor  <- as.factor(ToothGrowth$dose)
#   allFactor   <- as.factor(paste0(ToothGrowth$supp, "-", ToothGrowth$dose))
#   
#   ToothGrowth <- cbind(ToothGrowth, doseFactor)
#   ToothGrowth <- cbind(ToothGrowth, allFactor)
#   # print("All factor")
#   # print(allFactor)
#   # print(ToothGrowth)
# 
# createTeethData <- function ()
# {
#   teethDF <- data.frame(matrix(0, nrow = 10, ncol = 0))
#   # toothLen <- ToothGrowth$len
#   treatmentName <- c()
#   # print("ToothGrowth:")
#   # print(class(ToothGrowth))
#   # print(ToothGrowth)
#   # print("toothLen:")
#   # print(class(toothLen))
#   # print(toothLen)
#   # print(toothLen[1:10])
#   for (i in 1:nBins){
#     # print(ToothGrowth[((i-1)*nLength + 1) : (i*nLength),])
#     # print(cbind(ToothGrowth[((i-1)*nLength + 1) : (i*nLength),]))
#     # print(paste0(i, " start = ", (i-1)*nLength + 1, " end = ", i*nLength))
#     # print(toothLen[((i-1)*nLength + 1) : (i*nLength)])
#     # print("------------------------------------")
#     # # teethDF <- cbind(teethDF, toothLen[((i-1)*nLength + 1) : (i*nLength)])
#     # treatmentName[i] <- paste0(ToothGrowth$supp[(i - 1)*10 + 1], 
#     #                            "_",  ToothGrowth$dose[(i - 1)*10 + 1])
#   }
#   # colnames(ToothGrowth) <- treatmentName
#   # print(ToothGrowth)
#   # print(toothLen)
#   # print(teethDF)
#   
#   return (teethDF)

# }
```

## Images
### Convergence image

<!-- # ```{r smGraph, echo = FALSE, results = 'hide'} -->
```{r smGraph, echo = FALSE, fig.dim = c(4, 2)}

# createConvergencePlot <- function()
# {
  # setSource()
  # sampleStats  <- generateSampleStats(n, meansVector)
  # meansVector   <- setSource(sampleRepetition)
  # sampleStats  <- generateSampleStats(n, meansVector)
  # 
  # print("generateSampleStats: sampleStats")
  # print(sampleStats)

#   statDF <- cbind(statDF, variance = statDF$Variance + 5)
# 
#   print("createConvergencePlot:statDF")
#   print(statDF)
# 
#   graphTics   <- c(5, 10, 100, 1000, 10000, 100000, 1000000, 10000000)
# 
#   plt <- ggplot(statDF, aes(x = Size)) +
#     geom_line(aes(y=Mean, color='Mean')) +
#     geom_line(aes(y=Variance, color='Variance')) +
#     geom_hline(yintercept = 5.625, color="1/lambda") +
#     geom_hline(yintercept = 5, color="1/lambda") +
#     scale_x_log10(breaks = graphTics, limits = c(sampleRepetition[1],
#                   sampleRepetition[28]+1)) +
#     scale_y_continuous(limits = c( 4.0, 6.0 ), "Mean",
#     sec.axis = sec_axis(~ . - 5.0, name = "Variance")) +
#     ggtitle("Sample Mean and Variance") + xlab("Sample Size (log n)") +
#     scale_color_manual(
#       name='Statistic', breaks=c('Mean', 'Variance',"1/lambda"),
#       values=c('Mean'='blue', 'Variance'='red', "1/lambda"='black')) +
#     theme(plot.title = element_text(size = 10, hjust = 0.5),
#           axis.title.x = element_text(size = 8),
#           axis.title.y = element_text(size = 8),
#           axis.text.x = element_text(face="bold", size=6),
#           axis.text.y = element_text(face="bold", size=6),
#           legend.title = element_text(size = 6),
#           legend.text = element_text(size = 6),
#           legend.key.size = unit(2, 'mm'),
#           legend.position = c(0.85, 0.75))
# 
#   ggsave(plot = plt, width = 4.0, height = 2.5, dpi = 300,
#          filename = "converge.png")
# 
#   
# }
# 
# if(EXAMPLE & TEST_CONVERGE)
# {
#   colDistribution <- setSource()
#   createConvergencePlot()
# }

```

<!-- \begin{wrapfigure}{r}{0.3\textwidth} -->
<!--   \centering -->
<!--     \includegraphics[width=\linewidth]{converge.png} -->
<!-- \end{wrapfigure} -->

### teeth image

<!-- , results = 'hide' -->
```{r echo = FALSE, warning = FALSE, fig.dim = c(4, 3)}
# 
# createDotPlot <- function (x_axis)
# {
#   ToothGrowth$dotColor <- c(rep("blue",10),
#                             rep("green",10),
#                             rep("orange",10),
#                             rep("purple",10),
#                             rep("yellow",10),
#                             rep("red",10))
# 
#   g <- ggplot(ToothGrowth, aes_string(x = x_axis, y = "len"))
#   g + geom_boxplot() +
#     geom_dotplot(binaxis='y',
#                  stackdir='center',
#                  dotsize = .5,
#                  fill = ToothGrowth$dotColor) +
#     labs(title="Effect of Vitamin C on Tooth Growth",
#          subtitle="Delivery and Dosage",
#          caption="Source: ???",
#          x="Delivery (OJ, VC), Dosage(0.5, 1.0, 2.0)",
#          y="Tooth Length")
# 
# }

```


```{r echo = FALSE}

# createConvergencePlot()
# createHistograms()
# createDotPlot("allFactor")

```




<!-- \begin{wrapfigure}{r}{0.3\textwidth} -->
<!--   \centering -->
<!--     \includegraphics[width=\linewidth]{teeth.png} -->
<!-- \end{wrapfigure} -->


## Tables
```{r, echo = FALSE}
# 
#   statData <- data.frame(n = c(sampleStats$size[1],
#                           sampleStats$size[10],
#                           sampleStats$size[19],
#                           sampleStats$size[28]),
#                  Mean = c(sampleStats$mean[1],
#                           sampleStats$mean[10],
#                           sampleStats$mean[19],
#                           sampleStats$mean[28]),
#                  Deviation = c(sampleStats$st_dev[1],
#                                sampleStats$st_dev[10],
#                                sampleStats$st_dev[19],
#                                sampleStats$st_dev[28]))

# summarise_all(c("len", "dose"), mean)


```

```{r, echo = FALSE}
#
#   histogramData <-
#     data.frame(
#       n          = c(n[1], n[10],
#                      n[19], n[21]),
#       Mean       = c(round(sampleMean[[1]]),
#                          round(sampleMean[[2]]),
#                          round(sampleMean[[3]]),
#                          round(sampleMean[[4]])),
#       DeltaMu    = c("- - - -",
#                          round(sampleMean[[2]] - sampleMean[[1]], digits = 3),
#                          round(sampleMean[[3]] - sampleMean[[2]], digits = 3),
#                          round(sampleMean[[4]] - sampleMean[[3]], digits = 3)),
#       Sigma      = c(sampleDeviation[[1]],
#                          sampleDeviation[[2]],
#                          sampleDeviation[[3]],
#                          sampleDeviation[[4]]),
#       DeltaSigma = c("- - - -",
#                      round(sampleDeviation[[2]] - sampleDeviation[[1]],
#                            digits = 3),
#                      round(sampleDeviation[[3]] - sampleDeviation[[2]],
#                            digits = 3),
#                      round(sampleDeviation[[4]] - sampleDeviation[[3]],
#                            digits = 3))
#       )

```

### normal test
```{r echo = FALSE}

  # sampleMean      <- list()
  # sampleDeviation <- list()
  # comparisonSet <- c(1, 3, 10, 28)
  #
  # for (i in 1:length(comparisonSet)) {
  #   sampleMean[[i]]      <- mean(colDistributions[,comparisonSet[[i]]])
  #   sampleDeviation[[i]] <- sd(colDistributions[,comparisonSet[[i]]])
  # }

```

<!-- # Preliminaries -->
<!-- **Exponential Distribution** -->

```{r junk, echo=FALSE, results = 'hide'}

# setSource <- function ()
# {
#   for (i in 1:length(sampleRepetition)) {
#     # sampleMatrix <- matrix(rexp(sampleRepetition[i]*sample, lambda), 
#     #                                nrow = sampleRepetition[i], ncol = sample)
#     sampleMatrix <- matrix(rexp(sampleRepetition[i]*sample, lambda), 
#                                    nrow = sample, ncol = sampleRepetition[i])
# 
#     if((EXAMPLE & TEST_SOURCE) | FALSE)
#     {
#       filename <- paste0("sampleMatrix", sampleRepetition[i], ".csv")
#       write.csv(sampleMatrix, filename, row.names=FALSE)
#     }
# 
#     colDistribution   <- colMeans(sampleMatrix)
#     xbar      <- mean(colDistribution)
#     colVar    <- var(colDistribution)
# 
#     pdfList[i]  <<- list(colDistribution)
#     statDF      <<- rbind(statDF, c(sampleRepetition[i], xbar, colVar))
# 
#     if((EXAMPLE & TEST_SOURCE) | FALSE)
#     {
#       print("colDistribution")
#       # print(colDistribution)
#   
#       filename <- paste0("colDistribution", sampleRepetition[i], ".csv")
#       write.csv(colDistribution, filename, row.names=FALSE)
#   
#   
#       # colVar    <- 0
#       # for (j in 1:length(colDistribution)) {
#       #   colVar <- colVar + (colDistribution[j] - xbar)^2
#       # }
#       # colVar <- colVar/(sampleRepetition[i] - 1) # / sqrt(sampleRepetition[i])
#   
#       print(paste0("xbar = ", xbar, " colVar = ", colVar))
#     }
#   }
# 
#   colnames(statDF) <<- c("Size", "Mean", "Variance")
#   if((EXAMPLE & TEST_SOURCE) | FALSE)
#   {
#     print("class(colDistribution)")
#     print(is.matrix(colDistribution))
#     print(length(colDistribution))
# 
#     print("statDF out")
#     print(statDF)
#     # print("pdfList ")
#     # print(pdfList)
#   }
#   
#   print("setSource: statDF out")
#   print(statDF)
# 
#   return (colDistribution)
# }
# 
# 

### Generate sample stats

# ```{r generateSampleStats, echo=FALSE}
# generateSampleStats <- function (colDistribution)
# {
#   print("generateSampleStats: colDistribution")
#   # print(dim(colDistribution))
#   # print(colDistribution)
# 
#   sampleStats         <- data.frame()
#   for (i in 1:length(sampleRepetition)) {
#     # print("generateSampleStats: length(meansVector[i,]")
#     # print(length(meansVector[i,]))
#     # print("generateSampleStats: meansVector[i]")
#     # print(meansVector)
#     sampleStats  <-
#       rbind(sampleStats, list(sampleRepetition[i], mean(meansVector[i,]), var(meansVector[i,])))
#   }
#   colnames(sampleStats) <- c("Size", "Mean", "Variance")
# 
#   print("generateSampleStats: sampleStats")
#   # print(sampleStats)
#   filename <- paste0("sampleStats", n[i], ".csv")
#   write.csv(sampleStats, filename, row.names = FALSE)
# 
#   return (sampleStats)
# }
# # meansVector   <- setSource(sampleRepetition)
# # sampleStats  <- generateSampleStats(n, meansVector)
# 
# 
# if(EXAMPLE & SET_SOURCE)
# {
#   colDistribution <- setSource()
#   generateSampleStats(colDistribution)
# }


```
